Graph Topological Aspects of Granger Causal Network Learning

### 2019-07-14 paper/paper.tex
Trimmed down theory to create a draft for an application paper.

The following is a fundamental building block:

\begin{lemma}
  \label{lem:ancestor_uncorrelated}
  Consider distinct nodes $i, j$ in a Granger-causality graph
  $\gcg$.  Then, if

@@

  then $H_t^{(i)} \perp \H_t^{(j)}$, that is,
  $\forall \tau \in \Z_+\ \E[x_i(t)x_j(t - \tau)] = 0$.  Moreover,
  this implies that $j \npwgc i$.
\end{lemma}

\begin{remark}
  Note that $H_t^{(i)} \perp \H_t^{(j)}$ states that the \textit{present}
  of $x_i(t)$ is uncorrelated with the whole history of $x_j(t)$, but
  we need not have $\H_t^{(i)} \perp \H_t^{(j)}$ since $i \in \anc{j}$ is
  not excluded.
\end{remark}

\begin{proof}
  We show directly that
  $\forall \tau \in \Z_+\ \E[x_i(t)x_j(t - \tau)] = 0$.  To this end,
  fix $\tau \ge 0$, then by expanding with equation \eqref{eqn:parent_expansion} we have

  \begin{align*}
    \E x_i(t)x_j(t - \tau) &= \E \big(\A_{ii}(z)v_i(t)\big)\big(\A_{jj}(z)v_j(t - \tau)\big)\\
    &+ \sum_{\substack{k \in \anc{i} \\ k \ne i}}\E[\big(\A_{ik}(z)v_k(t)\big)v_j(t - \tau)] + \sum_{\substack{k \in \anc{j} \\ k \ne j}}\E[v_i(t) \big(\A_{jk}(z) v_k(t - \tau)\big)]\\
    &+ \sum_{\substack{k \in \anc{i} \\ k \ne i}}\sum_{\substack{\ell \in \anc{j} \\ \ell \ne j}}\E[\big(\A_{ik}(z)v_k(t)\big)\big(\A_{j\ell}(z)v_\ell(t - \tau)\big)].
  \end{align*}

% (i.e.
%   $\forall \tau \in \Z,\ k \ne \ell \Rightarrow \E v_k(t)v_\ell(t -
%   \tau) = 0$ and
%   $\forall \tau \in \Z\setminus\{0\}\ \E v_i(t)v_i(t - \tau) = 0$),

  Keeping in mind that $v(t)$ is an isotropic and uncorrelated
  sequence we will see that each of these above four terms are 0.  The
  first term since $i \ne j$.  The second since $j \not\in \anc{i}$.
  The fourth since $\anc{i} \cap \anc{j} = \emptyset$.  For the third
  term, it is possible that $i \in \anc{j}$, which requires more care.

  Similarly as before we can see immediately that

  \begin{equation*}
    \sum_{\substack{k \in \anc{j} \\ k \ne j}}\E[v_i(t) \big(\A_{jk}(z) v_k(t - \tau)\big)] = \E[v_i(t)\big(\A_{ji}(z)v_i(t - \tau)\big)].
  \end{equation*}

  However, $k \ne \ell \Rightarrow A_{k\ell}(0) = 0$ (i.e.
  $\A(z) = I + A(1)z^{-1} + \cdots$) so $\A_{ji}(z)v_i(t - \tau)$ is
  strictly earlier in time (for $\tau \ge 0$) than $v_i(t)$ and hence
  $\E[v_i(t)\big(\A_{ji}(z)v_i(t - \tau)\big)] = 0$.
\end{proof}

>>>>

\begin{lemma}
  \label{lem:separated_ancestor_uncorrelated}
  Consider distinct nodes $i, j$ in a Granger-causality graph
  $\gcg$.  If

  \begin{enumerate}[label=(\alph*)]
    \item{$j \not\in \anc{i}$ and $i \not\in \anc{j}$}
    \item{$\anc{i}\cap\anc{j} = \emptyset$}
  \end{enumerate}

  then $\H_t^{(i)} \perp \H_t^{(j)}$, that is,
  $\forall s, \tau \in \Z_+\ \E[x_i(t - s)x_j(t - \tau)] = 0$.  Moreover,
  this means that $j \npwgc i$ and $\linE{x_j(t)}{\H_{t - 1}^i} = 0$.
\end{lemma}

\begin{remark}
  The possibility that there exist nodes sharing both $i$ and $j$ as
  ancestors is not excluded, the point being that the temporal nature
  of Granger-causality eliminates the problems caused by ``colliders''
  (in the language of Pearl's causal calculus
  \cite{pearl2009causality}) in static causal inference.
\end{remark}

\begin{proof}
  We show directly that
  $\forall s, \tau \in \Z_+\ \E[x_i(t - s)x_j(t - \tau)] = 0$.  To this end,
  fix $s, \tau \ge 0$, then by expanding with equation \eqref{eqn:ancestor_expansion} we have

  \begin{align*}
    \E x_i(t - s)x_j(t - \tau)
    &= \E \big(\A_{ii}(z)v_i(t - s)\big)\big(\A_{jj}(z)v_j(t - \tau)\big)\\
    &+ \sum_{\substack{k \in \anc{i} \\ k \ne i}}\E[\big(\A_{ik}(z)v_k(t - s)\big)\big(\A_{jj}(z)v_j(t - \tau)\big)]\\
    &+ \sum_{\substack{k \in \anc{j} \\ k \ne j}}\E[\big(\A_{ii}(z)v_i(t - s)\big) \big(\A_{jk}(z) v_k(t - \tau)\big)]\\
    &+ \sum_{\substack{k \in \anc{i} \\ k \ne i}}\sum_{\substack{\ell \in \anc{j} \\ \ell \ne j}}\E[\big(\A_{ik}(z)v_k(t - s)\big)\big(\A_{j\ell}(z)v_\ell(t - \tau)\big)].
  \end{align*}

  Keeping in mind that $v(t)$ is an isotropic and uncorrelated
  sequence we see that each of these above four terms are 0: the
  first term since $i \ne j$, the second and third since
  $j \not\in \anc{i}$ and $i \not\in \anc{j}$ and finally the fourth since
  $\anc{i} \cap \anc{j} = \emptyset$.
\end{proof}

It is possible for components of $x(t)$ to be correlated at some time
lags without resulting in pairwise causality:

\begin{lemma}
  \label{lem:ancestor_uncorrelated}
  Consider distinct nodes $i, j$ in a Granger-causality graph $\gcg$.
  If

@@

  then $j \npwgc i$.
\end{lemma}
\begin{proof}
  By Theorem \ref{thm:granger_causality_equivalences} it suffices to show that

  \begin{equation*}
    \inner{x_i(t) - \linE{x_i(t)}{\H_{t - 1}^i}}{x_j(t) - \linE{x_j(t)}{\H_{t - 1}^i}} = 0,
  \end{equation*}

  which by the orthogonality principle is equivalent to

  \begin{equation}
    \label{eqn:proof_inner0}
    \inner{x_i(t)}{x_j(t) - \linE{x_j(t)}{\H_{t - 1}^i}} = 0.
  \end{equation}

  Define the disjoint sets

  \begin{align*}
    C_0(u) &= \{k \in \pa{u}\ |\ i \not\in \anc{k}, k \ne i \}\\
    C_1(u) &= \{k \in \pa{u}\ |\ i \in \anc{k} \text{ or } k = i \}.
  \end{align*}

  We can then expand the parents of $j$ using Equation \ref{eqn:parent_expansion} as

  \begin{equation*}
    x_j(t) = v_j(t) + \sum_{k \in C_0(j)} B_{jk}(z) x_k(t) + \sum_{k \in C_1(j)} B_{jk}(z) x_k(t),
  \end{equation*}

  which when substituted into the left hand side of Equation
  \ref{eqn:proof_inner0} results (by Lemma
  \ref{lem:separated_ancestor_uncorrelated}) in

  \begin{equation*}
    \inner{x_i(t)}{\sum_{k \in C_1(j)} B_{jk}(z) x_k(t) - \sum_{k \in C_1(j)} \linE{B_{jk}(z) x_k(t)}{\H_{t - 1}^i}}.
  \end{equation*}

  We can continue this process recursively (i.e. split each
  $k \in C_1(j)$ into $C_0(k)$ and $C_1(k)$) which must eventually
  terminate with $C_1(u) = \{i\}$.  Therefore, there exists some
  causal filter $\Phi(z)$ such that Equation \ref{eqn:proof_inner0} is
  equivalent to

  \begin{equation}
    \label{eqn:proof_inner0}
    \inner{x_i(t)}{\Phi(z)x_i(t) - \linE{\Phi(z)x_i(t)}{\H_{t - 1}^i}},
  \end{equation}

  which is $0$ since $\linE{\Phi(z)x_i(t)}{\H_{t - 1}^i} = \Phi(z)x_i(t)$.
\end{proof}

### 2019-07-14 paper_theory/paper_theory.tex
Trimmed down theory to create a draft for an application paper.

\documentclass[12pt]{article}

@@

%% bibliography stuff -- this needs come before the preamble inclusion
\usepackage[backend=bibtex,sorting=none]{biblatex}
\usepackage{enumitem}
\usepackage{etex,etoolbox}
\usepackage{hyperref}
\usepackage{fullpage}
\setlength{\marginparwidth}{2cm}
\usepackage{todonotes}
\bibliography{\string~/Documents/academics/global_academics/global_bib.bib}
% \usepackage{apxproof}

% \bibliography{\string~global_bib.bib}

@@

\input{\string~/Documents/academics/global_academics/latex_preamble}
% \input{\string~latex_preamble}

@@

\title{Graph Topological Aspects of Granger Causal Network Learning }
\author{R. J. Kinnear \\
  \small\href{mailto:ryan@kinnear.ca}{ryan@kinnear.ca} \\
  \small\url{https://github.com/RJTK} \and R. R. Mazumdar \\
  \small\href{mailto:mazum@uwaterloo.ca}{mazum@uwaterloo.ca} }

\begin{document}
\maketitle
\abstract{We study Granger-causality in the context of wide-sense
  stationary time series, where our focus is on the topological
  aspects of the underlying causality graph.  We establish sufficient
  conditions (in particular, we develop the notion of a ``strongly
  causal'' graph topology) under which the true causality graph can be
  recovered via pairwise causality testing alone, and provide examples
  from the gene regulatory network literature suggesting that our
  concept of a strongly causal graph may be applicable to this field.}

@@

% \tableofcontents
% \clearpage

@@

This notion is of interest for a variety of reasons.  From a purely
conceptual standpoint, we will see how the notion can in some sense
capture the idea of ``flow of information'' in the underlying graph,
in the sense that if $j \in \anc{i}$ we expect that $j \pwgc i$.  It may
also be useful for reasoning about the conditions under which
\textit{unobserved} components of $x(t)$ may or may not interfere with
inference in the actually observed components.  Finally, motivated
from a practical standpoint to analyze causation in large systems,
practical estimation procedures based purely on pairwise causality
tests are of interest since the computation of such pairwise relations
is substantially easier..

@@

It seems reasonable to expect a converse of proposition
\ref{prop:ancestor_properties} to hold, i.e.
$j \in \anc{i} \Rightarrow j \pwgc i$.  Unfortunately, this is not the
case in general, as different paths through $\gcg$ can lead to
cancellation (see example \ref{ex:diamond_cancellation}).  In fact, we
do not even have $j \in \pa{i} \Rightarrow j \pwgc i$ (see example
\ref{ex:lag_cancellation}).

\begin{example}
  \label{ex:diamond_cancellation}
  Firstly, on $n = 4$ nodes, ``diamond'' shapes can lead to cancellation on paths of length 2:

\begin{equation*}
  x(t) =
  \left[
    \begin{array}{cccc}
      0 & 0 & 0 & 0\\
      a & 0 & 0 & 0\\
      -a & 0 & 0 & 0\\
      0 & 1 & 1 & 0\\
    \end{array}
  \right] x(t - 1) + v(t),
\end{equation*}

with $\E v(t) = 0,\ \E v(t)v(t - \tau)^\T = \delta_\tau I$.

By directly calculating

\begin{align*}
  x_4(t) &= x_2(t - 1) + x_3(t - 1) + v_4(t)\\
         &= ax_1(t - 2) + av_2(t - 1) - ax_1(t - 2) -av_3(t - 1) + v_4(t)\\
         &= a(v_2(t - 1) - v_3(t - 1)) + v_4(t),
\end{align*}

we see that, since $v(t)$ is isotropic white noise, $1 \npwgc 4$.  The problem here is that there are multiple paths from $x_1$ to $x_4$.
\end{example}

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
    \includegraphics[width=\linewidth]{example1.pdf}
    \caption{Graph Corresponding to Example \ref{ex:diamond_cancellation}}
    \label{fig:diamond_cancellation}
    \centering{$j \in \anc{i} \nRightarrow j \pwgc i$}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
    \includegraphics[width=\linewidth]{example2.pdf}
    \caption{Graph Corresponding to Example \ref{ex:lag_cancellation}}
    \label{fig:lag_cancellation}
    \centering{$j \in \pa{i} \nRightarrow j \pwgc i$}
  \end{subfigure}
\end{figure}

\begin{example}
  \label{ex:lag_cancellation}
  A second example on $n = 3$ nodes is also worth examining, in this case
  cancellation is a result of differing time lags.

\begin{equation*}
  x(t) =
  \left[
    \begin{array}{ccc}
      0 & 0 & 0\\
      -a & 0 & 0\\
      0 & 1 & 0\\
    \end{array}
  \right] x(t - 1) +
  \left[
    \begin{array}{ccc}
      0 & 0 & 0\\
      0 & 0 & 0\\
      a & 0 & 0\\
    \end{array}
  \right] x(t - 2) + v(t)
\end{equation*}

Then

\begin{align*}
  x_2(t) &= v_2(t) - ax_1(t - 1)\\
  x_3(t) &= v_3(t) + x_2(t - 1) + ax_1(t - 2)\\
  \Rightarrow x_3(t) &= v_2(t - 1) + v_3(t),
\end{align*}

and again $1 \npwgc 3$.
\end{example}

@@

\clearpage
\printbibliography
\end{document}

>>>>

\documentclass{statsoc}
% \usepackage{mathtime}
\usepackage{ccfonts,eulervm}
% \usepackage[T1]{fontenc}
% \usepackage[lite]{belleek}
% \fontfamily{belleek}
% \usepackage[math]{belleek}
% \usepackage{concretemath-fonts}

% ---------------------------------------------------------------------------------------------------------------------------------
% The statsoc style file is broken?
% I have copied this code straight from stack overflow
% https://tex.stackexchange.com/questions/351768/problem-with-statsoc-class-and-pdflatex
\usepackage{geometry}

\geometry{
  textwidth=33pc,
  textheight=\dimexpr48\baselineskip+\topskip\relax,
  marginparsep=11pt,
  marginparwidth=107pt,
  footnotesep=6.65pt,
  headheight=9pt,
  headsep=9pt,
  footskip=30pt,
}
% ---------------------------------------------------------------------------------------------------------------------------------

% \usepackage{natbib}
\bibliographystyle{chicago}

@@

%% bibliography stuff -- this needs come before the preamble inclusion
\usepackage{enumitem}
\usepackage{etex,etoolbox}
\usepackage{hyperref}
\usepackage{fullpage}
\setlength{\marginparwidth}{2cm}
\usepackage{todonotes}

@@

% \usepackage{fullpage}
\usepackage{framed}

% Figures
\usepackage{graphicx}
\usepackage{caption}  % This is not recommended?
\usepackage{subcaption}
% \usepackage{wrapfig}
% \usepackage{svg}

% Math packages, theorem definitions and numbering
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{amsthm}
% \usepackage{mathrsfs} % Fancy scripted font
% \usepackage{bm}  % Bold math

% Misc packages
\usepackage[linesnumbered, ruled, vlined]{algorithm2e}
\usepackage{algorithm2e} %{algorithm} environment
\usepackage{soul}  % \hl highlighting
\usepackage{color}
\usepackage{mathtools}  % For my \ceil function

% Theorems (with italics)
% \theoremstyle{plain}  % Style definition removes italics
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}

% \theoremstyle{definition}
\newtheorem{remark}{Remark}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{assumptions}{Assumptions}

% keywords
\providecommand{\keywords}[1]{\textbf{\textit{Keywords---}} #1}

% General
\def\defeq{\overset{\Delta}{=}}  % Equal with triangle
\def\cl{\mathsf{cl\ }}  % Closure
\newcommand{\sgn}[1]{\mathsf{sgn}(#1)}  % sign function

% Calculus
\def\d{\mathsf{d}}  % Differential operator

% Functions
\def\ln{\mathsf{ln\ }}  % Natural logarithm
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}  % Ceiling

% Probability
\def\H{\mathcal{H}}  % Hilbert space
\def\E{\mathbb{E}}  % Expectation
\def\Var{\text{Var}}  % Variance
\def\P{\mathbb{P}}  % Probability Measure
\def\F{\mathcal{F}}  % A sigma algebra
\def\sX{\mathcal{X}}  % Another sigma algebra
\def\KL{\mathbf{D}_{KL}}  % KL divergence
\def\bF{\mathbf{F}}  % Whole F-meas space

% Standard sets
\def\Z{\mathbb{Z}}  % Set of integers
\def\R{\mathbb{R}}  % Set of real numbers
\def\C{\mathbb{C}}  % Set of complex numbers
\def\N{\mathbb{N}}  % Set of natural numbers
\def\ball{\mathbb{B}}  % Open ball
\def\clball{\overline{\ball}}  % Closed ball

% Linear algebra
\def\rk{\mathsf{rk }}  % The rank
\def\tr{\mathsf{tr }}  % The trace
\def\T{\mathsf{T}}  % Transpose notation
\def\c{\mathsf{c}}  % complement
\def\dg{\mathsf{dg }}   %  Diagonal vector of a matrix
\def\Dg{\mathsf{Dg }}   %  Diagonal matrix from a vector
\def\ind{\mathbf{1}}  % Ones vector or indicator
\def\matvec{\textbf{vec}}  % Vector operator
\def\<{\langle}  % < Inner product
\def\>{\rangle}  % > Inner product
\newcommand{\inner}[2]{\langle #1, #2 \rangle}  % Inner product
\newcommand{\innerT}[2]{#1^\T #2}  % Inner product for finite vectors

@@

\title{Graph Topological Aspects of Granger Causal Network Learning}
\author[Author 1 {\it et al.}]{R. J. Kinnear}
\address{
  University of Waterloo,
  Waterloo,
  Canada.}
\email{Ryan@Kinnear.ca}

\author{R. R. Mazumdar}
\address{
  University of Waterloo,
  Waterloo,
  Canada.}

\begin{document}

\abstract{We study Granger-causality in the context of wide-sense
  stationary time series, where our focus is on the topological
  aspects of the underlying causality graph.  We establish sufficient
  conditions (in particular, we develop the notion of a ``strongly
  causal'' graph topology) under which the true causality graph can be
  recovered via pairwise causality testing alone, and provide examples
  from the gene regulatory network literature suggesting that our
  concept of a strongly causal graph may be applicable to this field.}

@@

% \tableofcontents
\clearpage

@@

This notion is of interest for a variety of reasons.  From a purely
conceptual standpoint, we will see how the notion can in some sense
capture the idea of ``flow of information'' in the underlying graph,
in the sense that if $j \in \anc{i}$ we expect that $j \pwgc i$.  It may
also be useful for reasoning about the conditions under which
\textit{unobserved} components of $x(t)$ may or may not interfere with
inference in the actually observed components.  Finally, motivated
from a practical standpoint to analyze causation in large systems,
practical estimation procedures based purely on pairwise causality
tests are of interest since the computation of such pairwise relations
is substantially easier.

@@

It seems reasonable to expect a converse of proposition
\ref{prop:ancestor_properties} to hold, i.e.
$j \in \anc{i} \Rightarrow j \pwgc i$.  Unfortunately, this is not the
case in general, as different paths through $\gcg$ can lead to
cancellation (see Figure \ref{fig:diamond_cancellation}).  In fact, we
do not even have $j \in \pa{i} \Rightarrow j \pwgc i$ (see Figure
\ref{fig:lag_cancellation}).

% \begin{example}
%   \label{ex:diamond_cancellation}
%   Firstly, on $n = 4$ nodes, ``diamond'' shapes can lead to cancellation on paths of length 2:

% \begin{equation*}
%   x(t) =
%   \left[
%     \begin{array}{cccc}
%       0 & 0 & 0 & 0\\
%       a & 0 & 0 & 0\\
%       -a & 0 & 0 & 0\\
%       0 & 1 & 1 & 0\\
%     \end{array}
%   \right] x(t - 1) + v(t),
% \end{equation*}

% with $\E v(t) = 0,\ \E v(t)v(t - \tau)^\T = \delta_\tau I$.

% By directly calculating

% \begin{align*}
%   x_4(t) &= x_2(t - 1) + x_3(t - 1) + v_4(t)\\
%          &= ax_1(t - 2) + av_2(t - 1) - ax_1(t - 2) -av_3(t - 1) + v_4(t)\\
%          &= a(v_2(t - 1) - v_3(t - 1)) + v_4(t),
% \end{align*}

% we see that, since $v(t)$ is isotropic white noise, $1 \npwgc 4$.  The problem here is that there are multiple paths from $x_1$ to $x_4$.
% \end{example}

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.35\textwidth}
    \includegraphics[width=\linewidth]{example1.pdf}
    \caption{Example multiple path cancellation: $j \in \anc{i} \nRightarrow j \pwgc i$}
    \label{fig:diamond_cancellation}
  \end{subfigure}
  \begin{subfigure}[b]{0.35\textwidth}
    \includegraphics[width=\linewidth]{example2.pdf}
    \caption{Example cancellation from time lag: $j \in \pa{i} \nRightarrow j \pwgc i$}
    \label{fig:lag_cancellation}
  \end{subfigure}
\end{figure}

% \begin{example}
%   \label{ex:lag_cancellation}
%   A second example on $n = 3$ nodes is also worth examining, in this case
%   cancellation is a result of differing time lags.

% \begin{equation*}
%   x(t) =
%   \left[
%     \begin{array}{ccc}
%       0 & 0 & 0\\
%       -a & 0 & 0\\
%       0 & 1 & 0\\
%     \end{array}
%   \right] x(t - 1) +
%   \left[
%     \begin{array}{ccc}
%       0 & 0 & 0\\
%       0 & 0 & 0\\
%       a & 0 & 0\\
%     \end{array}
%   \right] x(t - 2) + v(t)
% \end{equation*}

% Then

% \begin{align*}
%   x_2(t) &= v_2(t) - ax_1(t - 1)\\
%   x_3(t) &= v_3(t) + x_2(t - 1) + ax_1(t - 2)\\
%   \Rightarrow x_3(t) &= v_2(t - 1) + v_3(t),
% \end{align*}

% and again $1 \npwgc 3$.
% \end{example}

@@

% \bibliography{\string~global_bib.bib}
\bibliography{\string~/Documents/academics/global_academics/global_bib}
\end{document}
